---
title: "Sample Analysis"
author: "Nermin Sinclair"
date: "10/13/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load library}
library(readr) # readr is in tidyverse
library(tidyverse)
library(dbplyr)
library(caret)
library(glmnet)
library(car)
```

```{r load file and define missing values}
# define missing values
na_values = c(" ", "NA", "Na", "na", "NaN")
CrimeDataChicago <- read_csv('CrimeData2007_07.csv', na= na_values)
# delete Total_CR
Transactiondata <- read_csv('Transactionstocommunities.csv', na= na_values) # deleted columns from source
```

```{r delete missing values}
# test for missing values
any(is.na(CrimeDataChicago))
any(is.na(Transactiondata))

# delete missing values
#CCrimeDataChicago <- na.omit(CrimeDataChicago)
Transactiondata  <- na.omit(Transactiondata)
any(is.na(Transactiondata))
```

```{r merge files}
# NOTE this only works because we have sample data will not work with full dataset
combinded = merge(x = Transactiondata, y = CrimeDataChicago, by.x ="COMMUNITY", by.y = "COMMUNITY")
any(is.na(combinded))
combinded = na.omit(combinded)
```

```{r delte columns with only one value, causes error in model}
# Delete Date and Transactio column because it has only one value causes error
combinded <- combinded[, -c(2,13)]
combinded <- combinded[, -6] #buyer state
combinded <- combinded[, -6] #buyer county
```

#### Prepare Cross-Validation and build train and test set
```{r Cross Validation with 10 folds}
fitControl <- trainControl(method = "cv", number= 10)
```

```{r create partition 85% training, 15% test WITH DATE}
# Y variable is dosage unit 
set.seed(2021)
partition <- createDataPartition(combinded$DOSAGE_UNI, p=.85)$Resample1
training_set<- combinded[partition,]
test_set <- combinded[-partition,]
```

#### Fit Model for Model Selection
```{r create/train glm poisson model}
# Generalized Linear Model Poisson Regression for Count Data
# Response Variable is a count
model_Poisson <- train(DOSAGE_UNI~ .,data=training_set, method = "glm", trControl=fitControl, family="poisson")
```

```{r}
model_Poisson
# look at Coefficients
model_Poisson$finalModel$coefficients
```
Look for similarities between the counties (to many counties have NA)

```{r}
combinded$COMMUNITY <- as.factor(combinded$COMMUNITY)
combinded$REPORTER_B <- as.factor(combinded$REPORTER_B)
combinded$REPORTER_S <- as.factor(combinded$REPORTER_S)
combinded$BUYER_BUS_ <- as.factor(combinded$BUYER_BUS_)
combinded$area_number <- as.factor(combinded$area_number)
```

plot(CovidDataCleaned_withDate)
par(mfrow=c(2,2))
plot(model_Poisson_withDate$finalModel) # doesn't look right!
# Detecting multicollinearity with Variance Inflation Factor
vif(model_Poisson_withDate$finalModel) # error message
# Plot pair-wise correlation among the explanatory variables
library(GGally)
counties <- CovidDataCleaned_withDate[, 2:57]
ggpairs(counties)                              #can't display graph
# Ferrar - Glauber Test
library(mctest)
imcdiag(x= counties, y= CovidDataCleaned_withDate$numConfC19Hosp) #mode missing, when added, still error about x
# partial correlation coefficient matrix 
library(corpcor)
cor2pcor(cov(counties))


```{r create elastic net model}
# Elastic net linear regression (lasso & ridge) to regularize regression model
# Variable selection and regularization simultaneously
model_ElasticNet <- train(DOSAGE_UNI~ ., data=training_set, method="glmnet", trControl=fitControl, family="poisson")
model_ElasticNet
coef(model_ElasticNet$finalModel, model_ElasticNet$bestTune$lambda)
```

```{r prediction on test set for validation}
pred_Poisson <- predict(model_Poisson, newdata=test_set)
pred_ElasticNet <- predict(model_ElasticNet, newdata = test_set)
# RMSE Model Peformance
data.frame(RMSE_Poisson = RMSE(pred_Poisson, test_set$DOSAGE_UNI), R2_Poisson= R2(pred_Poisson, test_set$DOSAGE_UNI),RMSE_EN = RMSE(pred_ElasticNet, test_set$DOSAGE_UNI), R2_EN= R2(pred_ElasticNet, test_set$DOSAGE_UNI))

```

